Evaluation of classical correlation functions from 2/3D images on CPU and GPU architectures: introducing CorrelationFunctions.jl

### 2023-09-15 paper-correlations.tex
Improvements from Alexey

\definecolor{light-gray}{gray}{0.95}
\newcommand{\code}[1]{\colorbox{light-gray}{\texttt{#1}}}
% When cleveref fails to do its job
\newcommand{\apref}[1]{Appendix \ref{#1}}

@@

\begin{abstract}
Correlation functions are becoming one of the major tools for quantification of
structural information that is usually represented as material's 2D or 3D image.
In this paper we introduce \code{CorrelationFunctions.jl} open-source package
developed in Julia and capable of computing all classical correlation functions
based on imaging input data. Images include both binary and multi-phase
representations.  Our code is capable of evaluating two-point probability $S_2$,
cluster $C_2$, lineal-path $L_2$, surface-surface $F_{ss}$, surface-void
$F_{sv}$, pore-size $P$ and chord-length $p$ functions on both CPU and GPU
architectures.  Where possible, we presented two types of computations: full
correlation map (correlations of each point with other points on the image, that
also allows obtaining ensemble averaged CF) and directional correlation
functions (currently in major orthogonal and diagonal directions).  Such an
implementation allowed for the first time to assemble a completely free solution
to evaluate correlation functions under any operating system with well
documented application programming interface (API). Our package includes
automatic tests against analytical solutions that are described in the paper. We
measured wall times for all CPU and GPU implementations and as a rule of thumb
full correlation maps on GPU are faster than other methods. However, full maps
require more RAM and, thus, are limited to available RAM resources. On the other
hand, directional CFs are memory efficient and can be evaluated for huge
datasets -- this way they are the first candidates for structural data
compression of feature extraction. The package itself is available through Julia
package ecosystem and on GitHub
(\url{https://github.com/fatimp/CorrelationFunctions.jl}), the latter source
also contains documentation and additional helpful resources such as tutorials.
We believe that a single powerful computational tool such as
\code{CorrelationFunctions.jl} will significantly facilitate the usage of
correlation functions in numerous areas of structural description and research,
as well as in machine learning applications. We also present some example as
applied to ceramic, soil composite and oil-bearing rock samples based on their
3D X-ray tomography and 2D scanning electron microscope images. Finally, we
conclude our paper with discussion of possible ways to further improve presented
computational framework.
\end{abstract}

@@

\section{Background and motivation}
\label{sec:background}
Materials are ubiquitous both in nature and in industrial human activities, and
their physical properties are interconnected to structure
\cite{Torquato_book,Sahimi_book}. The information about the structure usually
comes in the way of 2D and 3D images obtained by microscopy
\cite{moussaoui2018,neumann2019,FIB-SEMpaper}, tomography \cite{xctmat_book} and
other related methodologies.  This structural information can be used for
numerous types of analysis, including simulations of different processes within
studied materials or evaluation of their properties based on 3D structural data
\cite{youssef2005,Miao2017,FDMSS}. Unlike computational modelling techniques a
whole class of approximate methods exists including (rigorous) bounds that
allows very fast estimations of physical properties
\cite{eshelby1957,berryman1986use,rozanski2023} albeit the bounds themselves are
too wide to be applicable in the majority of practical problems. Nonetheless, it
is noteworthy that many of such bounds are originally based on structural
descriptors in the form of correlation functions (CFs).

@@

While evaluation of correlation functions from 2D and 3D images was a topic of
numerous research papers, the information is highly fragmented, especially in
terms of algorithms and their implementation in the code. Some snippets of code
are available for some functions, but they lack general API and are usually
implemented in proprietary interpreted languages such as Matlab. Due to a recent
leap forward in GPU-based computations, some papers describe computations on
this architecture, but the code is usually not provided. All in all, the
efficient and open-source single API solution utilizing both CPU and GPU
architectures is currently absent.

The aim of this paper is to establish a general open-source solution allowing to
compute all classical correlation functions. This solution should be well
documented with all computational algorithms explained, it should also leverage
on recent advances in programming high-intensity computations and utilize both
CPU and GPU power. Our answer in the form of \code{CorrelationFunctions.jl}
package allows to compute numerous CFs easily on any operating system. The
manuscript describing the package is organized as follows: in \cref{sec:math} we
introduce a reader to the most common and useful correlation functions,
\cref{sec:map} presents all algorithmic details for computing full CFs maps (map
method, full 2-point statistics for a given image). In \cref{sec:scan} we
discuss the second method to calculate correlation functions which is called the
scan approach. In \cref{sec:verification} we verify the all computations based on
known analytical solutions. \cref{sec:efficiency} deals with computational
efficiency of our code, we compare computational times for 2D and 3D images of
different sizes for scanning and map methods as evaluated using either CPU or
GPU. In \cref{sec:examples} some application examples are provided, including
CFs evaluation for multi-phase materials. Finally, \cref{sec:summary} summarizes
all results and outlines possible future improvements.

\section{A brief primer on correlation functions}
\label{sec:math}
In this section we shall provide major properties of correlation functions
supported by our package; the full list of functions is 
(see \cref{fig:functions} for graphical explanation in addition to definitions below):
\begin{itemize}
\item Two-point probability function $S_2$,
\item Lineal-path function $L_2$,
\item Cluster function $C_2$,
\item Surface-surface function $F_{ss}$,
\item Surface-void function $F_{sv}$,
\item Pore size function $P$,
\item Chord length function $p$,
\item Phase cross-correlation function $\rho_{ij}$.
\end{itemize}
Note that cross-correlation $\rho_{ij}$ is actually not a classical CF, but
technically is a variety of $S_2$ for the case when the end of line segment lie
in different phases. This function is useful for multi-phase structures only, as
for binary media two-point probability CFs are interdependent.

@@

Another basic correlation function is a lineal path function. It provides some
non-trivial information (albeit usually not complete \cite{vcapek2011transport})
about the connectedness of the phase and is defined for homogenous media as:
\begin{definition}
  Lineal path function $L_2^{(i)}(\bm{r})$ equals to probability that a
  vector $\bm{r}$ lies wholly in phase $i$ when randomly thrown into the
  sample.
\end{definition}
If a sample is isotropic a vector $\bm{r}$ can be replaced with its length
$r$ and the definition becomes:
\begin{definition}
  Lineal path function (for isotropic media) $L_2^{(i)}(r)$ equals to
  probability that a line segment of length $r$ lies wholly in phase $i$ when
  randomly thrown into the sample.
\end{definition}

@@

The next two functions are called surface-surface ($F_{ss}$) and surface-void
($F_{sv}$) functions and describe an interphase between phases. We skip strict verbal
definitions and first introduce an interface indicator function instead:
\begin{equation*}
  M^{(i)}(\bm{r}) = | \nabla I^{(i)}(\bm{r}) |
\end{equation*}
Here differentiation is understood in the sense of generalized functions,
because an ordinary differential of $I$ is either zero or undefined. For
homogenous media the definitions of surface functions then are:
\begin{align}
  F_{ss}^{(i)}(\bm{r}) &= \langle M^{(i)}(\bm{x}) M^{(i)}(\bm{x} +
  \bm{r}) \rangle \label{eq:fss-def} \\
  F_{sv}^{(i)}(\bm{r}) &= \langle M^{(i)}(\bm{x}) I^{(void)}(\bm{x}
  + \bm{r}) \rangle \label{eq:fsv-def}
\end{align}
These functions are defined for media with dimensionality of 2 and
higher. $F_{sv}$ is a physical quantity inversely proportional to length and
$F_{ss}$ is inversionaly proportional to surface area. As always, there are
versions of these functions applicable to scalar argument for isotropic media.

Pore size function $P(\delta)$ is a probability density function defined as
follows:
\begin{definition}
  $P(\delta)d\delta$ = Probability that a randomly chosen point in a set of points
  belonging to the void phase lies at a distance between $\delta$ and $\delta + d\delta$
  from the nearest point on the pore-solid interface.
\end{definition}

Finally, a chord length function $p^{(i)}(r)$ is defined as:
\begin{definition}
$p^{(i)}(r)dz$ = Probability of finding a chord of length between $r$ and $r+dr$
in phase $i$. Chord is a line segment which lies entirely in the same phase and
touches the interface with its ends.
\end{definition}

@@

\subsection{Full correlation map}
\label{sec:map}
Definition \cref{eq:s2-def} can be rewritten as follows:
\begin{equation}
  S_2^{(i)}(r) = \frac{F^{-1} [|F [A^{(i)}](z)|^2] (r)}{N(r)} \label{eq:s2-ft}
\end{equation}
where $A^{(i)} \xleftarrow{I^{(i)}} A$ is a binary array obtained by element-wise
application of interface indicator $I^{(i)}$ to an input $A$, $F$ is a discrete
Fourier transform (DFT) operator and $N(r)$ is a total number of trials (vectors
thrown into the sample). If we want to calculate $S_2$ function in periodic mode
we need only to apply \cref{eq:s2-ft} to an input to obtain the result. In
non-periodic mode we need to pad an input with zeros to at least twice the size
minus one for each dimension. In periodic mode a total number of trials is
independent of $r$ and equals to the number of elements in an input array. In
non-periodic mode $N(r)$ is calculated as described in
\cref{sec:number-of-trials}. The following algorithm summarizes all the above:
\begin{algorithmic}[1]
  \Procedure{$S_2$}{$A, phase, periodic$}
  \If{$\neg periodic$}
  \State $A \gets zeropad(A)$
  \Comment Pad with at least $2\cdot size(A, i) - 1$ zeros for each dimension $i$.
  \EndIf
  \State $A^{(phase)} \gets I^{(phase)} (A)$
  \State $\hat{A}^{(phase)} \gets F[A^{(phase)}]$
  \State $\hat{S}_2 \gets |\hat{A}^{(phase)}|^2 / N$
  \Comment $N$ is an array with numbers of trials
  \State \textbf{return} $F^{-1} [\hat{S}_2]$
  \EndProcedure
\end{algorithmic}
Algorithmic complexity of this implementation is $O(M \log M)$, where $M$ is the
number of elements in the input image.

@@

Cluster function can be calculated with the help of previuously described
procedures.  Firstly, we apply an indicator function to an input $A$:
$A^{(i)} \xleftarrow{I^{(i)}} A$. When we label connected components in
$A^{(i)}$ using a function that is a part of Julia \code{ImageMorphology.jl}
library.  Labeling algorithm must take into account if the function is
calculated with periodic boundary conditions \cite{evstigneev2023}. For periodic
boundary conditions we implemented our own solution as part of the package, as
it is absent in the public domain. Then $C_2$ function is a sum of $S_2$
functions calculated for all clusters separately. Algorithmic complexity of this
implementation is $O(MC\log M)$ where $M$ is the number of elements in the input
and $C$ is the number of clusters. The algorithm is then written as:
\begin{algorithmic}[1]
  \Procedure{$C_2$}{$A, phase, periodic$}
  \State $A^{(phase)} \gets I^{(phase)} (A)$
  \State $CC \gets label\_components(A^{(phase)}, periodic)$
  \State $N \gets maximum(CC)$ \Comment Number of connected components
  \State \textbf{return} $\sum\limits_{n=1}^N S_2(CC, n, periodic)$
  \EndProcedure
\end{algorithmic}

For surface-surface and surface-void functions we use an edge detection filter
\cite{samarin2023robust} and then calculate either autocorrelation of the edge
or cross-correlation of the edge and the void phase. The edge is extracted by
convolving an input with a short high-pass filter $H$. It has a width $7$ and
all its coefficients with exception of the central coefficient are inversely
proportional to distance from the center:
\begin{equation}
  \begin{aligned}
    H_k &= S \left\{
    \begin{array}{ll}
      -\sum\limits_{\substack{l \\ l \ne 0}} H_l & \quad k = 0 \\
      1 / \rho(k, 0) & \quad \text{otherwise}
    \end{array}
    \right. \\
    k &\in \{-3, -2, \dots, 3\}^D
  \end{aligned}
  \label{eq:filter-7x7}
\end{equation}
where $S$ equals to $30.45849$ in 2D and $172.96232$ in 3D case. The parameters
of the filter were adjusted based on comparison against exact computations of
surface functions on smooth-boundary sets \cite{postcov23}.

@@

Majority of described algorithms contain parallelizable operations such as fast
Fourier transform, connected components labeling and image filtering, hence they
are perfectly suited for execution on GPUs and provide very fast computations
(see wall times below).

@@

An algorithm for computation of linear-path function for one-dimensional slice
is presented below. Here $A$ is a one-dimensional input array, $phase$ can be
void, solid or some other phase, and $periodic$ is a boolean value which equals
to true if we compute $L_2$ in periodic mode and false otherwise. The efficiency
of the proposed algorithm is $O(n)$ where $n$ is the length of the array.
\begin{algorithmic}[1]
  \Procedure{l2}{$A, phase, periodic$}
    \State $len \gets length(A)$
    \State $result \gets zeros(len)$
    \State $runs \gets countruns(A, phase)$
    \ForAll{$run \in runs$}
      \State $updateruns(result, run)$
    \EndFor
    \If{$periodic$}
      \State $first \gets first(A)$
      \State $last \gets last(A)$
      \If{$first = last$}
        \State $updateperiodic(result, first, last)$
      \EndIf
    \EndIf
    \State \textbf{return} $success$ divided by number of trials
    (see \cref{sec:number-of-trials})
  \EndProcedure
  \\
  \Procedure{countruns}{$A, phase$}
    \State $runslist \gets [\ ]$
    \State $runs \gets 0$
    \ForAll{$x \in A$}
      \If{$x = phase$}
        \State $runs \gets runs + 1$
      \ElsIf{$runs \ne 0$}
        \Comment Add $runs$ to the accumulator
        \State $runslist \gets runs:runslist$
        \State $runs \gets 0$
      \EndIf
    \EndFor
    \If{$runs \ne 0$}
      \State $runslist \gets runs:runslist$
    \EndIf
    \State \textbf{return} $runslist$
  \EndProcedure
  \\
  \Procedure{updateruns}{$A, runs$}
    \State $r \gets runs$
    \For{$idx \gets 0, runs - 1$}
      \State $A[idx] \gets A[idx] + r$
      \State $r \gets r - 1$
    \EndFor
  \EndProcedure
  \\
  \Procedure{updateperiodic}{$A, first, last$}
    \State $sum \gets first + last$
    \For{$idx \gets 0, sum - 1$}
      \State $up \gets \min(idx, first, last, sum - idx)$
      \State $A[idx] \gets A[idx] + up$
    \EndFor
  \EndProcedure
\end{algorithmic}

Cluster function can be calculated using the same technique as in map approach,
but if the number of clusters is high, a naïve $O(n^2)$ algorithm may be
preferred to $O(MC \log M)$ full map algorithm. Such an approach is present below:
\begin{algorithmic}[1]
  \Procedure{c2}{$A, phase, periodic$}
    \State $A' \gets I^{(phase)}(A)$
    \State $CC \gets label\_components(A', periodic)$
    \State $len \gets length(A)$
    \State $result \gets zeros(len)$
    \ForAll{$slice \in slices(CC)$}
      \For{$i = 0,len-1$}
        \State $end \gets len-1$
        \If{$periodic$}
          \State $end \gets end + i$
        \EndIf
        \For{$j = i, end$}
          \If{$slice[i] = slice[j \mod len] \ne 0$}
            \Comment $0$ is a label for background.
            \State $dst \gets |(j \mod len) - i|$
            \State $result[dst] \gets result[dst] + 1$
          \EndIf
        \EndFor
      \EndFor
    \EndFor
    \State \textbf{return} $result$ divided by number of trials
  \EndProcedure
\end{algorithmic}

Before computation of chord length function an input array must be pre-processed
to find an interface for the phase of interest. For this purpose we use
distance transform function defined as follows:
\begin{equation}
  \mathcal{D}(\bm{x})= \left\{
  \begin{array}{ll}
    0 & \quad \bm{x} \in \text{solid phase} \\
    \min\limits_{y \in \text{solid phase}} \rho(\bm{x},\bm{y}) & \quad \text{otherwise}
  \end{array}
\right. \label{eq:distance-transform}
\end{equation}
There are many algorithms for distance transform, including algorithms performed
in linear time \cite{DT}.In \code{chord\_length} function we firstly apply
distance transform to an input to find the interface and then accumulate chord
length in per-slice manner. Finally, chord lengths can be binned to a histogram.
\begin{algorithmic}[1]
  \Procedure{chord\_length}{$A, phase$}
    \State $A' \gets I^{(phase)}(A)$
    \State $A_E \gets \mathcal{D}(A') = 1$
    \Comment We consider elements for which the distance transform equals to $1$
    to be part of the interface
    \State $chord\_list \gets [\ ]$
    \ForAll{$(S', S_E) \in zip(slices(A'), slices(A_E))$}
      \Comment iterate through all one-dimensional slices in $A'$ and
      corresponding slices in $A_E$
      \State $l \gets 0$
      \State $maybe\_chord \gets False$
      \For{$idx \gets 0, length(S')-1$}
        \If{$S_E[idx]$}
          \Comment Interface found
          \If{$l > 1\ and\ maybe\_chord$}
            \Comment Chord found
            \State $chord\_list \gets l:chord\_list$
          \EndIf
          \State $l \gets 0$
          \State $maybe\_chord \gets True$
        \ElsIf{$\neg S'[idx]$}
          \Comment We are not in the phase of interest
          \State $maybe\_chord \gets False$
        \EndIf
        \State $l \gets l+1$
      \EndFor
    \EndFor
    \State \textbf{return} $chord\_list$ or $histogram(chord\_list)$
  \EndProcedure
\end{algorithmic}
This implementation also works in linear time.

>>>>

\definecolor{light-gray}{gray}{0.95}
\newcommand{\code}[1]{\colorbox{light-gray}{\texttt{#1}}}

@@

\begin{abstract}
Correlation functions are becoming one of the major tools for quantification of
structural information that is usually represented as 2D or 3D image of porous
material.  In this paper we introduce \code{CorrelationFunctions.jl} open-source
package developed in Julia and capable of computing all classical correlation
functions based on imaging input data. Images include both binary and
multi-phase representations.  Our code is capable of evaluating two-point
probability $S_2$, phase cross-correlation $\rho_{ij}$, cluster $C_2$,
lineal-path $L_2$, surface-surface $F_{ss}$, surface-void $F_{sv}$, pore-size
$P$ and chord-length $p$ distribution functions on both CPU and GPU
architectures.  Where possible, we presented two types of computations: full
correlation map (correlations of each point with other points on the image, that
also allows obtaining ensemble averaged CF) and directional correlation
functions (currently in major orthogonal and diagonal directions).  Such an
implementation allowed for the first time to assemble a completely free solution
to evaluate correlation functions under any operating system with well
documented application programming interface (API). Our package includes
automatic tests against analytical solutions that are described in the paper. We
measured execution times for all CPU and GPU implementations and as a rule of
thumb full correlation maps on GPU are faster than other methods. However, full
maps require more RAM and, thus, are limited to available RAM resources. On the
other hand, directional CFs are memory efficient and can be evaluated for huge
datasets -- this way they are the first candidates for structural data
compression of feature extraction. The package itself is available through Julia
package ecosystem and on GitHub
(\url{https://github.com/fatimp/CorrelationFunctions.jl}), the latter source
also contains documentation and additional helpful resources such as tutorials.
We believe that a single powerful computational tool such as
\code{CorrelationFunctions.jl} presented in this paper will significantly
facilitate the usage of correlation functions in numerous areas of structural
description and research of porous materials, as well as in machine learning
applications. We also present some example as applied to ceramic, soil composite
and oil-bearing rock samples based on their 3D X-ray tomography and 2D scanning
electron microscope images. Finally, we conclude our paper with discussion of
possible ways to further improve presented computational framework.
\end{abstract}

@@

\section{Background and motivation}
\label{sec:background}
Heterogeneous materials are ubiquitous both in nature and in industrial human
activities, and their physical properties are interconnected to structure
\cite{Torquato_book,Sahimi_book}. The information about the structure usually
comes in the way of 2D and 3D images obtained by microscopy
\cite{moussaoui2018,neumann2019,FIB-SEMpaper}, tomography \cite{xctmat_book} and
other related methodologies.  This structural information can be used for
numerous types of analysis, including simulations of different processes within
studied materials or evaluation of their properties based on 3D structural data
\cite{youssef2005,Miao2017,FDMSS}. Unlike computational modelling techniques a
whole class of approximate methods exists including (rigorous) bounds that
allows very fast estimations of physical properties
\cite{eshelby1957,berryman1986use,rozanski2023} albeit the bounds themselves are
too wide to be applicable in the majority of practical problems. Nonetheless, it
is noteworthy that many of such bounds are originally based on structural
descriptors in the form of correlation functions (CFs).

@@

While evaluation of correlation functions from 2D and 3D images was a topic of
numerous research papers, the information is highly fragmented, especially in
terms of algorithms and their implementation in the code. Some snippets of code
are available for some functions, but they lack general API and are usually
implemented in proprietary interpreted languages (e.g. Matlab). Due to a recent
leap forward in GPU-based computations, some papers describe computations on
this architecture, but the code is usually not provided. All in all, the
efficient and open-source single API solution utilizing both CPU and GPU
architectures is currently absent.

The aim of this paper is to establish a general open-source solution allowing to
compute all classical correlation functions. This solution should be well
documented with all computational algorithms explained, it should also leverage
on recent advances in programming high-intensity computations and utilize both
CPU and GPU power. Our answer in the form of \code{CorrelationFunctions.jl}
package allows to compute numerous CFs easily on any operating system. The
manuscript describing the package is organized as follows: in \cref{sec:math} we
introduce a reader to the most common and useful correlation functions,
\cref{sec:map} presents all algorithmic details for computing full CFs maps (map
method, full 2-point statistics for a given image). In \cref{sec:scan} we
discuss the second method to calculate correlation functions which is called the
scan approach. In \cref{sec:verification} we verify our algorithms based on
known analytical solutions. In \cref{sec:examples} some application examples are
provided, including CFs evaluation for multi-phase materials. Section
\ref{sec:efficiency} deals with computational efficiency of our code, we compare
computational times for 2D and 3D images of different sizes for scanning and map
methods as evaluated using either CPU or GPU.  Finally, sections
\ref{sec:outlook} and \ref{sec:summary} summarize all results and outline
possible future improvements.

\section{A brief primer on correlation functions}
\label{sec:math}
In this section we shall provide major properties and definitions of correlation
functions supported by our package; the full list of functions is (see
\cref{fig:functions} for graphical explanation in addition to definitions
below):
\begin{itemize}
\item Two-point probability function $S_2$,
\item Lineal-path function $L_2$,
\item Cluster function $C_2$,
\item Surface-surface function $F_{ss}$,
\item Surface-void function $F_{sv}$,
\item Pore size function $P$,
\item Chord length function $p$,
\item Phase cross-correlation function $\rho_{ij}$.
\end{itemize}
Note that cross-correlation $\rho_{ij}$ is actually not a classical CF, but
technically is a variety of $S_2$ for the case when the end of line segment lie
in different phases. This function is useful for multi-phase structures only, as
for binary media two-point probability CFs are interdependent.

@@

Another basic correlation function is a lineal path function $L_2$. It provides
some non-trivial information (albeit usually not complete
\cite{vcapek2011transport}) about the connectedness of the phase and is defined
for homogenous media as:
\begin{definition}
  Lineal path function $L_2^{(i)}(\bm{r})$ equals to probability that a
  vector $\bm{r}$ lies wholly in phase $i$ when randomly thrown into the
  sample.
\end{definition}
If a sample is isotropic a vector $\bm{r}$ can be replaced with its length
$r$ and the definition becomes:
\begin{definition}
  Lineal path function (for isotropic media) $L_2^{(i)}(r)$ equals to
  probability that a line segment of length $r$ lies wholly in phase $i$ when
  randomly thrown into the sample.
\end{definition}

@@

The next two functions are called surface-surface ($F_{ss}$) and surface-void
($F_{sv}$) correlation functions and describe an interphase between phases. We
skip strict verbal definitions and first introduce an interface indicator
function instead:
\begin{equation*}
  M^{(i)}(\bm{r}) = | \nabla I^{(i)}(\bm{r}) |
\end{equation*}
Here differentiation is understood in the sense of generalized functions,
because an ordinary differential of $I$ is either zero or undefined. For
homogenous media the definitions of surface functions then are:
\begin{align}
  F_{ss}^{(i)}(\bm{r}) &= \langle M^{(i)}(\bm{x}) M^{(i)}(\bm{x} +
  \bm{r}) \rangle \label{eq:fss-def} \\
  F_{sv}^{(i)}(\bm{r}) &= \langle M^{(i)}(\bm{x}) I^{(void)}(\bm{x}
  + \bm{r}) \rangle \label{eq:fsv-def}
\end{align}
These functions are defined for media with dimensionality of 2 and
higher. $F_{sv}$ is a physical quantity inversely proportional to length and
$F_{ss}$ is inversionaly proportional to surface area. As always, there are
versions of these functions applicable to scalar argument for isotropic media.

Pore size function $P(r)$ is a probability density function defined as follows:
\begin{definition}
  $P(r)dr$ equals to the probability that a randomly chosen point in a set of
  points belonging to the void phase lies at a distance between $r$ and $r + dr$
  from the nearest point on the pore-solid interface.
\end{definition}

Finally, a chord length function $p^{(i)}(r)$ is defined as:
\begin{definition}
  $p^{(i)}(r)dr$ equals to the probability of finding a chord of length between
  $r$ and $r+dr$ in phase $i$. Chord is a line segment which lies entirely in
  the same phase and touches the interface with its ends.
\end{definition}

@@

\subsection{Full correlation map}
\label{sec:map}
Definition \cref{eq:s2-def} can be rewritten as follows:
\begin{equation}
  S_2^{(i)}(r) = \frac{F^{-1} [|F [A^{(i)}](z)|^2] (r)}{N(r)} \label{eq:s2-ft}
\end{equation}
where $A^{(i)} \xleftarrow{I^{(i)}} A$ is a binary array obtained by element-wise
application of indicator function $I^{(i)}$ to an input $A$, $F$ is a discrete
Fourier transform (DFT) operator and $N(r)$ is a total number of trials (vectors
thrown into the sample). If we want to calculate $S_2$ function in periodic mode
we need only to apply \cref{eq:s2-ft} to an input to obtain the result. In
non-periodic mode we need to pad an input with zeros to at least twice the size
minus one for each dimension. In periodic mode a total number of trials is
independent of $r$ and equals to the number of elements in an input array. In
non-periodic mode $N(r)$ is calculated as described in
\cref{sec:number-of-trials}. The following algorithm summarizes all the above:
\begin{algorithmic}[1]
  \Procedure{$S_2$}{$A, phase, periodic$}
  \If{$\neg periodic$}
  \State $A \gets zeropad(A)$
  \Comment Pad with at least $2\cdot size(A, i) - 1$ zeros for each dimension $i$.
  \EndIf
  \State $A^{(phase)} \gets I^{(phase)} (A)$
  \State $\hat{A}^{(phase)} \gets F[A^{(phase)}]$
  \State $\hat{S}_2 \gets |\hat{A}^{(phase)}|^2 / N$
  \Comment $N$ is an array with numbers of trials
  \State \textbf{return} $F^{-1} [\hat{S}_2]$
  \EndProcedure
\end{algorithmic}
Algorithmic complexity of this implementation is $O(M \log M)$, where $M$ is the
number of elements in the input image.

@@

Cluster function can be calculated with the help of previuously described
procedures.  Firstly, we apply an indicator function to an input $A$:
$A^{(i)} \xleftarrow{I^{(i)}} A$. Then we label clusters in $A^{(i)}$ using a
function that is a part of Julia \code{ImageMorphology.jl} library.  Labeling
algorithm must take into account if the function is calculated with periodic
boundary conditions \cite{evstigneev2023}. For periodic boundary conditions we
implemented our own solution as part of the package, as it is absent in the
public domain. Then $C_2$ function is a sum of $S_2$ functions calculated for
all clusters separately. Algorithmic complexity of this implementation is
$O(MC\log M)$ where $M$ is the number of elements in the input and $C$ is the
number of clusters. The algorithm is then written as:
\begin{algorithmic}[1]
  \Procedure{$C_2$}{$A, phase, periodic$}
  \State $A^{(phase)} \gets I^{(phase)} (A)$
  \State $CC \gets label\_components(A^{(phase)}, periodic)$
  \State $N \gets maximum(CC)$ \Comment Number of connected components
  \State \textbf{return} $\sum\limits_{n=1}^N S_2(CC, n, periodic)$
  \EndProcedure
\end{algorithmic}

For surface-surface and surface-void functions we use an edge detection filter
\cite{samarin2023robust} and then calculate either autocorrelation of the edge
or cross-correlation of the edge and the void phase. The edge is extracted by
convolving an input with a short high-pass filter $H$. It has a width of 7
pixels/voxels and all its coefficients with exception of the central coefficient
are inversely proportional to distance from the center:
\begin{equation}
  \begin{aligned}
    H_k &= S \left\{
    \begin{array}{ll}
      -\sum\limits_{\substack{l \\ l \ne 0}} H_l & \quad k = 0 \\
      1 / \rho(k, 0) & \quad \text{otherwise}
    \end{array}
    \right. \\
    k &\in \{-3, -2, \dots, 3\}^D
  \end{aligned}
  \label{eq:filter-7x7}
\end{equation}
where $S$ equals to $30.45849$ in 2D and $172.96232$ in 3D case. The parameters
of the filter were adjusted based on comparison against exact computations of
surface functions on smooth-boundary sets \cite{postcov23}.

@@

Majority of described algorithms contain parallelizable operations such as fast
Fourier transform, connected components labeling and image filtering, hence they
are perfectly suited for execution on GPUs and provide very fast computations
(see execution times below).

@@

An algorithm for computation of linear-path function for one-dimensional slice
is presented below. Here $A$ is a one-dimensional input array, $phase$ can be
void, solid or some other phase, and $periodic$ is a boolean value which equals
to true if we compute $L_2$ in periodic mode and false otherwise. The efficiency
of the proposed algorithm is $O(n)$ where $n$ is the length of the array.
\begin{algorithmic}[1]
  \Procedure{$L_2$}{$A, phase, periodic$}
    \State $len \gets length(A)$
    \State $result \gets zeros(len)$
    \State $runs \gets countruns(A, phase)$
    \ForAll{$run \in runs$}
      \State $updateruns(result, run)$
    \EndFor
    \If{$periodic$}
      \State $first \gets first(A)$
      \State $last \gets last(A)$
      \If{$first = last$}
        \State $updateperiodic(result, first, last)$
      \EndIf
    \EndIf
    \State \textbf{return} $success$ divided by number of trials
    (see \cref{sec:number-of-trials})
  \EndProcedure
  \\
  \Procedure{countruns}{$A, phase$}
    \State $runslist \gets [\ ]$
    \State $runs \gets 0$
    \ForAll{$x \in A$}
      \If{$x = phase$}
        \State $runs \gets runs + 1$
      \ElsIf{$runs \ne 0$}
        \Comment Add $runs$ to the accumulator
        \State $runslist \gets runs:runslist$
        \State $runs \gets 0$
      \EndIf
    \EndFor
    \If{$runs \ne 0$}
      \State $runslist \gets runs:runslist$
    \EndIf
    \State \textbf{return} $runslist$
  \EndProcedure
  \\
  \Procedure{updateruns}{$A, runs$}
    \State $r \gets runs$
    \For{$idx \gets 0, runs - 1$}
      \State $A[idx] \gets A[idx] + r$
      \State $r \gets r - 1$
    \EndFor
  \EndProcedure
  \\
  \Procedure{updateperiodic}{$A, first, last$}
    \State $sum \gets first + last$
    \For{$idx \gets 0, sum - 1$}
      \State $up \gets \min(idx, first, last, sum - idx)$
      \State $A[idx] \gets A[idx] + up$
    \EndFor
  \EndProcedure
\end{algorithmic}

Cluster function can be calculated using the same technique as in map approach,
but if the number of clusters is high, a naïve $O(n^2)$ algorithm may be
preferred to $O(MC \log M)$ full map algorithm. Such an approach is present below:
\begin{algorithmic}[1]
  \Procedure{$C_2$}{$A, phase, periodic$}
    \State $A' \gets I^{(phase)}(A)$
    \State $CC \gets label\_components(A', periodic)$
    \State $len \gets length(A)$
    \State $result \gets zeros(len)$
    \ForAll{$slice \in slices(CC)$}
      \For{$i = 0,len-1$}
        \State $end \gets len-1$
        \If{$periodic$}
          \State $end \gets end + i$
        \EndIf
        \For{$j = i, end$}
          \If{$slice[i] = slice[j \mod len] \ne 0$}
            \Comment $0$ is a label for background.
            \State $dst \gets |(j \mod len) - i|$
            \State $result[dst] \gets result[dst] + 1$
          \EndIf
        \EndFor
      \EndFor
    \EndFor
    \State \textbf{return} $result$ divided by number of trials
  \EndProcedure
\end{algorithmic}

Before computation of chord length function an input array must be pre-processed
to find an interface for the phase of interest. For this purpose we use
distance transform function defined as follows:
\begin{equation}
  \mathcal{D}(\bm{x})= \left\{
  \begin{array}{ll}
    0 & \quad \bm{x} \in \text{solid phase} \\
    \min\limits_{y \in \text{solid phase}} \rho(\bm{x},\bm{y}) & \quad \text{otherwise}
  \end{array}
\right. \label{eq:distance-transform}
\end{equation}
where $\rho$ is the Euclidean distance between two points. There are many
algorithms for distance transform, including algorithms performed in linear time
\cite{DT}.In \code{chord\_length} function we firstly apply distance transform
to an input to find the interface and then accumulate chord length in per-slice
manner. Finally, chord lengths can be binned to a histogram.
\begin{algorithmic}[1]
  \Procedure{chord\_length}{$A, phase$}
    \State $A' \gets I^{(phase)}(A)$
    \State $A_E \gets \mathcal{D}(A') = 1$
    \Comment We consider elements for which the distance transform equals to $1$
    to be part of the interface
    \State $chord\_list \gets [\ ]$
    \ForAll{$(S', S_E) \in zip(slices(A'), slices(A_E))$}
      \Comment iterate through all one-dimensional slices in $A'$ and
      corresponding slices in $A_E$
      \State $l \gets 0$
      \State $maybe\_chord \gets False$
      \For{$idx \gets 0, length(S')-1$}
        \If{$S_E[idx]$}
          \Comment Interface found
          \If{$l > 1\ and\ maybe\_chord$}
            \Comment Chord found
            \State $chord\_list \gets l:chord\_list$
          \EndIf
          \State $l \gets 0$
          \State $maybe\_chord \gets True$
        \ElsIf{$\neg S'[idx]$}
          \Comment We are not in the phase of interest
          \State $maybe\_chord \gets False$
        \EndIf
        \State $l \gets l+1$
      \EndFor
    \EndFor
    \State \textbf{return} $chord\_list$ or $histogram(chord\_list)$
  \EndProcedure
\end{algorithmic}
This implementation also works in linear time.
